---
title: "Structural differences"
author: "Ricardo Ramirez"
date: "9/8/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(cowplot)
library(mistyR)
library(uwot)
library(factoextra)
```

## Introduction

Can we explain the differences of phenotypes using cell-cell interactions?

The assumption is that the compositional differences observed at the single cell level should in one way or another affect the way cells organize and generate structures

First we recover the importances 

```{r message=F, warning=F}
misty_out_folder <- "./results/tissue_structure/misty/cell_map/"
misty_outs <- list.files(misty_out_folder, full.names = F)
misty_outs <- set_names(misty_outs, gsub("cm_", "", misty_outs) %>%
                          gsub("_c2l", "", .))

sample_dict <- readRDS("./markers/visium_patient_anns_revisions.rds")

# Here we will extract the importances of all views in all samples
sample_importances <- map(misty_outs, function(x) {
  misty_res <- collect_results(paste0(misty_out_folder, x))[["importances.aggregated"]]
})

sample_importances <- sample_importances %>% 
  enframe() %>% 
  unnest() %>%
  na.omit() %>%
  left_join(sample_dict, by = c("name" = "sample_id"))
```

## Transform to patient specific importances

```{r}
pat_importances <- sample_importances %>%
  group_by(view, Predictor, Target, patient_id) %>%
  summarise(Importance = mean(Importance))
```

## Generate the matrix for clustering and UMAP

First of all data

```{r}
pat_importances_mat <- pat_importances %>%
  ungroup() %>%
  #dplyr::filter(view == "intra") %>%
  dplyr::mutate(int_id = paste0(view,"..",Predictor,"_to_",Target)) %>%
  dplyr::select(patient_id, int_id, Importance) %>%
  pivot_wider(names_from = int_id, values_from = Importance) %>%
  column_to_rownames("patient_id") %>%
  as.matrix()

write_csv(as.data.frame(pat_importances_mat), 
          file = "./results/tissue_structure/colocalization/misty_importances_ct_bypat.csv")
```

Generate clustering as in other analysis

Ward with Eucledian

```{r}
imp_hclust <- eclust(pat_importances_mat, "hclust", k = 3)

color_palette <- tibble(patient_id = imp_hclust$labels[imp_hclust$order]) %>%
  left_join(sample_dict[,c("patient_group", "patient_id")] %>% unique()) %>%
  left_join(tibble(patient_group = c("group_1", "group_2", "group_3"),
                   col = c("red", "darkgreen", "blue")))

allviews_clust <- fviz_dend(imp_hclust, 
               rect = TRUE, 
               label_cols = color_palette$col,
               k_colors = rep("black",3),
               main = "all_views")

pdf("./results/tissue_structure/colocalization/misty_ct_importances.pdf", height = 7, width = 5)
plot(allviews_clust)
dev.off()

plot(allviews_clust)
```

Separate the analysis for each view

```{r}
view_pat_importances_mat <- pat_importances %>%
  ungroup() %>%
  group_by(view) %>%
  nest() %>%
  mutate(data = map(data, function(dat) {
    
    dat %>%
      dplyr::mutate(int_id = paste0(view,"..",Predictor,"_to_",Target)) %>%
      dplyr::select(patient_id, int_id, Importance) %>%
      pivot_wider(names_from = int_id, values_from = Importance) %>%
      column_to_rownames("patient_id") %>%
      as.matrix()
    
  }))
  #dplyr::filter(view == "intra") %>%
```

Cluster per view

```{r}
view_pat_importances_mat <- view_pat_importances_mat %>%
  mutate(cluster_tree = map2(view, data, function(v, dat) {
    
    
imp_hclust <- eclust(dat, "hclust", k = 3)

color_palette <- tibble(patient_id = imp_hclust$labels[imp_hclust$order]) %>%
  left_join(sample_dict[,c("patient_group", "patient_id")] %>% unique()) %>%
  left_join(tibble(patient_group = c("group_1", "group_2", "group_3"),
                   col = c("red", "darkgreen", "blue")))
fviz_dend(imp_hclust, 
               rect = TRUE, 
               label_cols = color_palette$col,
               k_colors = rep("black",3),type = "rectangle", main = v)
    
  }))

walk(view_pat_importances_mat$cluster_tree, plot)

pdf("./results/tissue_structure/colocalization/misty_ct_importances_clustbyview.pdf", height = 9, width = 5)

walk(view_pat_importances_mat$cluster_tree, plot)

dev.off()

```

We perform a PCA on the scaled importances

```{r}
ilr_pcs_res <- prcomp(x = pat_importances_mat)
ilr_pcs <- ilr_pcs_res$x %>%
  as.data.frame() %>%
  rownames_to_column("orig.ident")
```

```{r}
ilr_pcs %>%
left_join(sample_dict %>% 
            dplyr::select(patient_group, patient_id) %>%
            unique(), by = c("orig.ident" = "patient_id")) %>%
  ggplot(aes(x = PC1, y = PC2, 
             color = patient_group, 
             label = orig.ident)) +
      geom_point(size = 1) +
      ggrepel::geom_text_repel() +
      theme_classic()
```

This projection is misleading, since it only takes into account the first 2 PCs aprox 12%

```{r}
prop_sd = ilr_pcs_res$sdev/sum(ilr_pcs_res$sdev)
PC_label = paste0("PC_", seq(1,length(ilr_pcs_res$sdev),1))

tibble("prop_sd" = prop_sd,
       "PC" = PC_label) %>%
  ggplot(aes(x = factor(PC,
                        levels = PC_label), 
             y = prop_sd)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90))

```

## UMAP of samples

Finally, we project the pseudobulk profiles in the UMAP and define three groups of patients from the hierarchical clustering of the expression profiles of all samples

```{r}
gex_umap <- umap(pat_importances_mat, 
                 n_neighbors = 5, 
                 n_epochs = 1000,
                 metric = "cosine") %>%
  as.data.frame() %>%
  mutate(orig.ident = rownames(pat_importances_mat))
```

```{r, fig.height=5, fig.width=7}
umap_plt <- gex_umap %>%
  left_join(sample_dict %>% 
            dplyr::select(patient_group, patient_id) %>%
            unique(), by = c("orig.ident" = "patient_id")) %>%
  ggplot(aes(x = V1, y = V2, 
             color = patient_group, 
             label = orig.ident)) +
      geom_point(size = 1) +
      ggrepel::geom_text_repel() +
      theme_classic() +
  xlab("UMAP1") +
  ylab("UMAP2")

plot(umap_plt)
```

```{r}
gex_umap %>%
  left_join(sample_dict %>% 
            dplyr::select(patient_group, patient_id) %>%
            unique(), by = c("orig.ident" = "patient_id")) %>%
  write_csv(., file = "./results/sample_comparison/spatial/misty_importances_umap.csv")


pdf("./results/sample_comparison/spatial/misty_importances_umap.pdf",width = 5, height = 4.5)
plot(umap_plt)
dev.off()
```

Then we test the importances that are higher in one group vs the other:
We are correcting first by cell-cell importance 

## Get R performance

```{r}
misty_res <- collect_results(paste0(misty_out_folder, misty_outs))

R2_data <- misty_res$improvements %>%
  dplyr::filter(measure == "multi.R2") %>%
  dplyr::mutate(sample = gsub("_c2l", "", sample) %>%
                  strsplit(.,split = "cm_") %>%
                  map_chr(., ~ last(.x))) %>%
  dplyr::left_join(sample_dict, by = c("sample" = "sample_id")) %>%
  rename("R2" = value)

filt_predictors <- R2_data %>%
  dplyr::select(target, sample, R2) %>%
  dplyr::filter(R2 >= 5) %>%
  dplyr::mutate(keep = TRUE) %>%
  dplyr::select(target,R2, sample,keep)
```

## Filter 

```{r}
sample_importances_filt <- sample_importances %>%
  left_join(filt_predictors, by = c("Target" = "target",
                                    "name" = "sample")) %>%
  na.omit()
```


```{r, warning=F}
run_wilcox_all <- function(prop_data) {
  
  prop_data_group <- prop_data[["patient_group"]] %>%
    unique() %>%
    set_names()
  
  map(prop_data_group, function(g) {
    
    test_data <- prop_data %>%
      mutate(test_group = ifelse(patient_group == g,
                                 "target", "rest")) %>%
      mutate(test_group = factor(test_group,
                                 levels = c("target", "rest")))
    
    wilcox.test(Importance ~ test_group, 
                data = test_data,
                alternative = "greater") %>%
      broom::tidy()
  }) %>% enframe("patient_group") %>%
    unnest()
  
}

```

```{r}

# Here we identify interactions that are important ( > 1)
# We will only test view/predictor/target in which the 
# median of one group is larger than 1 (to reduce tests)

# You want to keep then all interactions that 
# were significant but at the same time that have a relevant importance in general

summarize_int_comparison <- function(corr_pval_filt, corr_pval_text, max_median_imp_filt =  1, 
                                     sample_importances,
                                     patient_group_filt = c("group_1", "group_2", "group_3")) {
  
  cell_interaction_comp <- sample_importances %>%
  dplyr::filter(patient_group %in% patient_group_filt) %>%
  group_by(view, Predictor, Target) %>%
  nest() %>%
  mutate(max_median_imp = map(data, function(dat) {
    
    dat %>% 
      group_by(patient_group) %>%
      summarize(med_importance = median(Importance)) %>%
      dplyr::filter(med_importance == max(med_importance)) %>%
      pull(med_importance)
    
  })) %>%
  unnest(max_median_imp) %>%
  dplyr::filter(max_median_imp >= max_median_imp_filt) %>%
  mutate(wres = map(data, run_wilcox_all)) %>%
  #select(wres) %>%
  unnest(wres) %>%
  ungroup() %>%
  group_by(view, Predictor) %>%
  mutate(corr_pval = p.adjust(p.value)) %>%
  ungroup()
  
  all_sign_ints <- cell_interaction_comp %>%
  dplyr::filter(corr_pval <= corr_pval_filt) %>%
  ungroup() %>%
  dplyr::mutate(int_id = paste0(Predictor, "_to_", Target)) %>%
  dplyr::select(view, data, Predictor, Target, patient_group, corr_pval, int_id, max_median_imp) %>%
  mutate(pval_text = ifelse(corr_pval <= corr_pval_text, "*", "")) %>%
  group_by(Predictor, Target) %>%
  mutate(ref_imp = max(max_median_imp)) %>%
  dplyr::filter(ref_imp > 0)

lab_order <- all_sign_ints %>%
  arrange(patient_group, int_id) %>%
  pull(int_id) %>% unique()

imprt_plot <- all_sign_ints %>%
  dplyr::select(-data) %>%
  mutate(int_id = factor(int_id,
                         levels = lab_order)) %>%
  #complete(view, int_id, patient_group) %>%
  ggplot(., aes(x = patient_group, 
                          y = int_id, 
                          fill = -log10(corr_pval))) +
  geom_tile() +
  geom_text(aes(label=pval_text)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.background = element_rect(fill="white", colour='black',
                                        size=1)) +
  facet_wrap(.~view, ncol = 3) +
  ylab("Model interaction") +
  scale_fill_gradient(na.value = "white")
 
 return(list("data" = all_sign_ints, "plot" = imprt_plot)) 
  
}
```

## Healthy and FZ

```{r}

healthy_vs_fz <- summarize_int_comparison(corr_pval_filt = 0.25, 
                                          corr_pval_text = 0.15,
                                          max_median_imp_filt = 1,
                                          sample_importances = sample_importances,
                                          patient_group_filt = c("group_1", "group_3"))

healthy_vs_iz <- summarize_int_comparison(corr_pval_filt = 0.25, 
                                          corr_pval_text = 0.15,
                                          max_median_imp_filt = 1,
                                          sample_importances = sample_importances,
                                          patient_group_filt = c("group_1", "group_2"))

fz_vs_iz <- summarize_int_comparison(corr_pval_filt = 0.25, 
                                          corr_pval_text = 0.15,
                                          max_median_imp_filt = 1,
                                          sample_importances = sample_importances,
                                          patient_group_filt = c("group_2", "group_3"))

all <- summarize_int_comparison(corr_pval_filt = 0.25, 
                                          corr_pval_text = 0.15,
                                          max_median_imp_filt = 1,
                                          sample_importances = sample_importances,
                                          patient_group_filt = c("group_1", "group_2", "group_3"))

```

```{r}
all_comps <- list("healthy_vs_fz" = healthy_vs_fz, 
     "healthy_vs_iz" = healthy_vs_iz,
     "fz_vs_iz" = fz_vs_iz, 
     "all" = all) %>%
  enframe()

walk2(all_comps$name, all_comps$value, function(tag, l) {
  
  pdf(paste0("./results/sample_comparison/spatial/sign_interaction_changes_",
             tag,".pdf"), height = 5, width = 5)
  
  plot(l[["plot"]])
  
  dev.off()
  
  write_csv(l[["data"]], paste0("./results/sample_comparison/spatial/sign_interaction_changes_",
             tag,".csv"))
  
})

write.csv(sample_importances, "./results/sample_comparison/spatial/sample_importances.csv")
```

```{r}
tags <- c("healthy_vs_fz", 
             "healthy_vs_iz",
             "fz_vs_iz")

all_interactions <- map(set_names(tags), ~ paste0("./results/sample_comparison/spatial/sign_interaction_changes_",
                      .x,
                      ".csv") %>%
      read_csv) %>%
  enframe() %>%
  unnest()


all_interactions <- all_interactions %>%
  dplyr::select(name, Predictor, Target) %>%
  pivot_longer(-name,names_to = "cell_type") %>%
  group_by(name, value) %>%
  summarise(ntimes = n()) %>%
  arrange(name, -ntimes) %>%
  mutate(prop_int = ntimes/sum(ntimes))


dplyr::filter(all_interactions, name == "healthy_vs_fz")

dplyr::filter(all_interactions, name == "healthy_vs_iz")

dplyr::filter(all_interactions, name == "fz_vs_iz")

```

```{r}

healthy_vs_fz_filt <- summarize_int_comparison(corr_pval_filt = 0.25, 
                                          corr_pval_text = 0.15,
                                          max_median_imp_filt = 0,
                                          sample_importances = sample_importances_filt,
                                          patient_group_filt = c("group_1", "group_3"))

healthy_vs_iz_filt <- summarize_int_comparison(corr_pval_filt = 0.25, 
                                          corr_pval_text = 0.15,
                                          max_median_imp_filt = 0,
                                          sample_importances = sample_importances_filt,
                                          patient_group_filt = c("group_1", "group_2"))

fz_vs_iz_filt <- summarize_int_comparison(corr_pval_filt = 0.25, 
                                          corr_pval_text = 0.15,
                                          max_median_imp_filt = 0,
                                          sample_importances = sample_importances_filt,
                                          patient_group_filt = c("group_2", "group_3"))

all_filt <- summarize_int_comparison(corr_pval_filt = 0.25, 
                                          corr_pval_text = 0.15,
                                          max_median_imp_filt = 0,
                                          sample_importances = sample_importances_filt,
                                          patient_group_filt = c("group_1", "group_2", "group_3"))

```


```{r}
all_comps <- list("healthy_vs_fz" = healthy_vs_fz_filt, 
     "healthy_vs_iz" = healthy_vs_iz_filt,
     "fz_vs_iz" = fz_vs_iz_filt, 
     "all" = all_filt) %>%
  enframe()

walk2(all_comps$name, all_comps$value, function(tag, l) {
  
  pdf(paste0("./results/sample_comparison/spatial/sign_interaction_changes_filt_",
             tag,".pdf"), height = 5, width = 5)
  
  plot(l[["plot"]])
  
  dev.off()
  
  write_csv(l[["data"]], paste0("./results/sample_comparison/spatial/sign_interaction_changes_filt_",
             tag,".csv"))
  
})

write.csv(sample_importances_filt, "./results/sample_comparison/spatial/sample_importances_filt.csv")
```

```{r}
tags <- c("healthy_vs_fz", 
             "healthy_vs_iz",
             "fz_vs_iz")

all_interactions <- map(set_names(tags), ~ paste0("./results/sample_comparison/spatial/sign_interaction_changes_filt_",
                      .x,
                      ".csv") %>%
      read_csv) %>%
  enframe() %>%
  unnest()


all_interactions <- all_interactions %>%
  dplyr::select(name, Predictor, Target) %>%
  pivot_longer(-name,names_to = "cell_type") %>%
  group_by(name, value) %>%
  summarise(ntimes = n()) %>%
  arrange(name, -ntimes) %>%
  mutate(prop_int = ntimes/sum(ntimes))


dplyr::filter(all_interactions, name == "healthy_vs_fz")

dplyr::filter(all_interactions, name == "healthy_vs_iz")

dplyr::filter(all_interactions, name == "fz_vs_iz")

```